### 2016-02-16 Projects/DelCo_2012/Report/DelCo_2012.tex
Fixing Typos in Body added text to Summary
The summary for the gas cooling is about 75% complete. Dan will
complete.

\begin{table}[!ht]
\centering
\captionof{table}{Tests to Assess Fire Fighting Agents Compared to Water}\label{tab:agent_tests}
\begin{tabular}{lcc}
\toprule[1.5pt]
Test                                    & Similar Performance & Enhanced Performance \\
\midrule
Specific Heat                           &  \checkmark         &                      \\
Fuel Cooling and Penetration            &                     &  \checkmark          \\
Mass Retention                          &                     &  \checkmark          \\
Ignition Inhibition                     &                     &  \checkmark          \\
Drop Size                               &  \checkmark         &                      \\
Contact Angle                           &  \checkmark         &                      \\
Tire Fire Suppression                   &  \checkmark         &                      \\
Wood Crib Suppression                   &  \checkmark         &                      \\
Heptane Fire Suppression                &  \checkmark         &                      \\
Magnesium \& Titanium Fire Suppression  &  \checkmark         &                      \\
Smoke Generation                        &  \checkmark         &                      \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

Based on the results of the study, the fire fighting agents, commonly referred to as Class A foams, were superior to water in cooling and penetrating the fuels, mass retention and ignition inhibition.  All of these features enable the agents, when applied in the form of foam, to provide enhanced radiation exposure protection to a structure relative to plain water.  The foams enhance the water by keeping the water in place on vertical surfaces such as walls and the reduced interfacial surface tension allows the water to penetrate absorbent fuels.  These enhancements also make the agents valuable tools for mop-up. Many of the experiments conducted during this study: specific heat, drop size, contact angle, wood crib suppression, smoke generation and Class D fire suppression yielded similar results for both Class A foams and plain water.

@@

The fuel load for these experiments consisted of 2 sleeper sofas and 2 chairs, carpet, padding, and wood paneling along the southeast corner walls. An image of the fuel arrangement is shown in Fig.~\ref{fig:Furniture_Fuel_Load}. The couches and chairs are setup to mimic a typical seating area with one couch against the south and east wall respectively with each being 0.9~m (36~in) away from the corner. A chair is placed adjecent to each couch and the specific locations can be found in the dimensioned schematic, Fig.~\ref{fig:Furniture_Fuel_Load_Dimensions}. In total, this fuel package contains approximately 336 kg of fuel consisting of wood, polyurethane (PU), polyester (PET),and polyprolene (PP). Table~\ref{tab:Fire_Suppression_Fuel_Masses} provides a breakdown on the fuel composition for these experiments.

@@

\chapter{Discussion and Tactical Considerations}
\label{chap:Discussion_and_Tactical_Considerations}

Water is the most widely used fire extinguishing agent because it is effective, environmentally friendly, nontoxic, inexpensive and in many cases, readily available. In addition, water has a very high heat of vaporization per unit mass, at least four times as high as that of any other nonflammable liquid~\cite{NFPA}. The latent heat of vaporization of water is 2255~kJ/kg (970~Btu/lb.)~\cite{NFPA}. This means that 2255~kJ (2139~Btu) of energy is required to change 1~kg (2.2~lb.) of water into steam.  When water is vaporized, its volume increases approximately 1,600 times. Because the energy absorbing capabilities of water are well quantified, they can be used as a basis to calculate the theoretical minimum delivery rate of water needed to extinguish a burning material with a known heat (energy) release rate. Unfortunately, experience has shown that water must be applied at 10 to 100 times the theoretical rate in practice to control and extinguish the fire~\cite{Friedman:2}. As a result of this apparent inefficiency and the need to address fires containing a wide variety of materials, water-based fire fighting additives have been utilized for many years to enhance the fire fighting capabilities of ordinary water.

@@

>>>>

\begin{table}[!ht]
\centering
\captionof{table}{Tests to Assess Fire Fighting Agents Compared to Water}\label{tab:agent_tests}
\begin{tabular}{lcc}
\toprule[1.5pt]
Test                                    & Similar Performance & Enhanced Performance \\
\midrule
Specific Heat                           &  \checkmark         &                      \\
Fuel Cooling and Penetration            &                     &  \checkmark          \\
Mass Retention                          &                     &  \checkmark          \\
Ignition Inhibition                     &                     &  \checkmark          \\
Drop Size                               &  \checkmark         &                      \\
Contact Angle                           &                     &  \checkmark          \\
Tire Fire Suppression                   &  \checkmark         &                      \\
Wood Crib Suppression                   &  \checkmark         &                      \\
Heptane Fire Suppression                &  \checkmark         &                      \\
Magnesium \& Titanium Fire Suppression  &  \checkmark         &                      \\
Smoke Generation                        &  \checkmark         &                      \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

Based on the results of the NIST 1998 study, the fire fighting agents, commonly referred to as Class A foams, were superior to water in reducing contact angle, cooling and penetrating the fuels, mass retention on the fuels and ignition inhibition of the fuels.  When applied in the form of foam, the water agent solutions provide enhanced radiation exposure protection to a structure relative to plain water.  The foams enhance the water by keeping the water in place on vertical surfaces such as walls and the reduced inter-facial surface tension allows the water to penetrate absorbent fuels.  These enhancements also make the agents valuable tools for mop-up. The other experiments conducted during this study: specific heat, drop size, wood crib suppression, smoke generation and Class D fire suppression yielded similar results for both Class A foams and plain water.

@@

The fuel load for these experiments consisted of 2 sleeper sofas and 2 chairs, carpet, padding, and wood paneling along the southeast corner walls. An image of the fuel arrangement is shown in Fig.~\ref{fig:Furniture_Fuel_Load}. The couches and chairs are setup to mimic a typical seating area with one couch against the south and east wall respectively with each being 0.9~m (36~in) away from the corner. A chair is placed adjacent to each couch and the specific locations can be found in the dimensioned schematic, Fig.~\ref{fig:Furniture_Fuel_Load_Dimensions}. In total, this fuel package contains approximately 336 kg of fuel consisting of wood, polyurethane (PU), polyester (PET),and polypropylene (PP). Table~\ref{tab:Fire_Suppression_Fuel_Masses} provides a breakdown on the fuel composition for these experiments.

@@

\chapter{Discussion}
\label{chap:Discussion}

This study focused on two key types of fire experiments; gas cooling and fire suppression. In each type of experiment a comparison was conducted between plain water and compressed air foam being discharged from nozzles of similar type with similar mass flow rates.  In both types of experiments there were no measurable differences between the use of plain water or compressed air foam. These results were not surprising to some of the workshop attendees.  However other workshop attendees, that use CAF very successfully in their fire departments, had additional questions. The research team worked to address those follow-up questions by conducting additional fire suppression experiments, one where only wood based fuels were used and another where a larger space with more ventilation was used.  This section will discuss the results and provide potential explanations to support them.

The gas cooling experiments were designed to measure the ability of water streams and CAF streams to cool a hot gas layer with a minimum temperature of  250~$^{\circ}$C (482~$^{\circ}$F) at positions 1.83\~{}m (6\~{}ft) below the ceiling.  During the planning of the experiments, some firefighters thought that the excellent heat absorbing capacity of plain water would exceed the heat absorbing capabilities of CAF, since the injected air, air being a great insulator, would interfere and result in less cooling. It was the belief that the reduced gas cooling capabilities of CAF may have lead to firefighter injuries during interior fire attacks.  Other firefighters thought that the bubble structure would provide the same mass flowrate of water with more surface area, which would enable a faster heat transfer rate and therefore, improve the cooling of the hot gas layer.  There were also thoughts that the CAF would coat the ceiling and eliminate energy feedback from the ceiling to the hot gas layer, thereby providing enhanced cooling.

Water is the most widely used fire extinguishing agent because it is effective, environmentally friendly, nontoxic, inexpensive and in many cases, readily available. In addition, water has a very high heat of vaporization per unit mass, at least four times as high as that of any other nonflammable liquid~\cite{NFPA}. The latent heat of vaporization of water is 2255~kJ/kg (970~Btu/lb.)~\cite{NFPA}. This means that 2255~kJ (2139~Btu) of energy is required to change 1~kg (2.2~lb.) of water into steam.  When water is vaporized, its volume increases approximately 1,700 times. Because the energy absorbing capabilities of water are well quantified, they can be used as a basis to calculate the theoretical minimum delivery rate of water needed to extinguish a burning material with a known heat (energy) release rate. Unfortunately, experience has shown that water must be applied at 10 to 100 times the theoretical rate in practice to control and extinguish the fire~\cite{Friedman:2}. As a result of this apparent inefficiency and the need to address fires containing a wide variety of materials, water-based fire fighting additives have been utilized for many years to enhance the fire fighting capabilities of ordinary water.

@@

In each gas cooling experiment, the same mass flow rate of water was used for both the plain water stream and the CAF stream.  Typically the volumetric flow rate of water was 120 gpm.  The density of water is 8.3 lbs/gal    therefore the mass flow rate was    .  The average time of flow was 15 s, therefore a total of X kg of water was introduced into each gas cooling experiment.  When a CAFs stream was used it contained 0.3\% foam concentration by volume, check density of foam concentrate   point being mostly water.  Also calc the mass of air in the stream.

Compare flow rate with coverage estimates Iowa/NFA formulas.

Ref Sean Mitchell report and then close with the ref back to the two comparison graphs.   Under the conditions of a residential scale room and contents fire it should be no surprise that the results are similar, the mass of water introduced into the hot gas layer is doing the work.  Air filled bubbles in the CAFs stream had no discernible impact in these experiments.
